File-based localization of numerical perturbations in data analysis pipelines

### 2020-08-12 paper.tex
Revise comments

\revised{
The labeled graph can be different depending on the order of executions in which condition 
we execute pipeline to pinpoint the propagation of differences. 
Therefore, we label processes in two order of executions, and then aggregate findings so that label processes 
as non-reproducible (red) if it creates different results in both order of executions.
}

\subsection{Implementation}
\revised{ 
The tool is implemented in Python programming language and supports the Python version 3.5 and higher.
It utilizes the Python modules including NumPy~\cite{oliphant2006guide} and Pandas~\cite{mckinney2011pandas}, 
which allow data processing, 
SciPy~\cite{2020SciPy-NMeth} and Scikit-learn~\cite{pedregosa2011scikit} packages 
for clustering the subjects, and Zss~\cite{zhang1989simple} package to calculate the tree distances.
Also, the Python APIs of Docker and Boutiques~\cite{glatard2018boutiques} tools were used to create the container 
images and simplify the pipeline executions.

We used the latest version of Boutiques tool v0.5.25 for creating descriptors 
to execute pipelines across the operating systems automatically.
In a JSON schema, the descriptors define inputs, outputs, error codes, containers, and the command line for running pipelines.

Besides, we used \reprozip tool version 1.0.16 to collect the pipeline system call processes and 
files in an SQLite file format.
}

@@

We randomly selected 20 unprocessed subjects from the HCP data release S500
available in \href{https://db.humanconnectome.org}{the ConnectomDB
repository} as a subset of the 1200 Subject Release. For each subject,
available data consisted of 1 or 2 T1-weighted images and 1 or 2
T2-weighted images, with $256 \times 320 \times 320$ voxels of size $0.7
\times 0.7 \times 0.7$ mm. Acquisition protocols and parameters are
detailed in~\cite{van2013wu}.

@@

% \revision{Using the \reprozip trace obtained from CentOS7, 
% we captured transient files in CentOS6 and labeled pipeline processes in CentOS7.}
We processed the 20 subjects with PreFreeSurfer and FreeSurfer, using the 2
CentOS versions. 
Each subject was processed twice on the same operating
system to detect within-OS variability coming from pseudo-random
operations. We compared pipeline results using  FreeSurfer tools \texttt{mri\_diff},
\texttt{mris\_diff}, and \texttt{lta\_diff}, to ignore execution-specific information such as file path or
timestamps. To compare segmentations $X$ and $Y$, we used the Dice coefficient defined as follows:
\[DICE=\frac{2|X \cap Y|}{|X| + |Y|}\]

\revised{
The Dice coefficient~\cite{dice1945measures} is a commonly used metric in validating medical image segmentation.
The Dice value ranges from 0 to 1, with 1 indicating a complete overlap between two segmentation results.
Besides, the Jaccard coefficient~\cite{jaccard1912distribution} can be considered an equivalent metric 
to measure the overlap between images.
}

@@

\section{Results}
\revised{
All experiments were run on a machine with a 3.4GHz Intel Core i7 processor and 32GB memory 
under the CentOS7 operating system. 
The details of the computation of the PreFreeSurfer and FreeSurfer pipelines are illustrated 
in~Table~\ref{table:pipeline-stats}.
The scripts and analyses used to create the figures in this section are available at 
\url{https://github.com/big-data-lab-team/HCP-reproducibility-paper}.

\begin{table*}[ht]
  \centering
  \begin{threeparttable}
  \caption{The computation details of the pipelines per subject.}
  \label{table:pipeline-stats}
  \begin{tabular}{ccc|cc}
  \toprule
  \multicolumn{1}{c}{} & \multicolumn{2}{c}{PreFreeSurfer } &   \multicolumn{2}{c}{FreeSurfer } \\

   &        \makecell{Mean}   &  \makecell{Standard error}  & \makecell{Mean}   &  \makecell{Standard error}   \\ \midrule
  Processing time (mins)      &     106.67     &     2.68     &     nd     &     nd     \\
  Output file size (GB)       &     2.8        &     0.10     &     nd     &     nd     \\
  Number of file accesses     &     94089      &     2644.99  &     nd     &     nd     \\
  Number of processes         &     8731       &     198.18   &     nd     &     nd     \\
  \bottomrule
  \end{tabular}
  \end{threeparttable}
  \end{table*}
}
% All experiments were run on a machine with a 3.4GHz Intel Core i7 processor 
% and 32GB memory under the CentOS7 operating system.
% The average processing time per subject was approximately 2~hours (with standard error X) 
% for PreFreeSurfer and 8~hours (with standard error X) for FreeSurfer. 
% The average output file size was 2.7~GB (with standard error Y) for PreFreeSurfer 
% and 4.1~GB (with standard error Y) for FreeSurfer.
%\tristan{Ali, are these numbers per subject or for 20 subjects? it's an average of 20 subjects}
% For each subject, PreFreeSurfer accessed 83K files and created 7.7K processes, and FreeSurfer accessed 62K files and 
% created 4K processes.

@@

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figures/t2w_alignment.png} 
    \caption{Differences between T2 \fnirt results in PreFreeSurfer's Brain Extraction (CentOS6 vs 
    CentOS7). 
    \revised{
    The colored squares indicate result images obtained from CentOS6 (in purple) and CentOS7 (in green), 
    and the red boxes highlight regions with significant differences.}
    % Result images obtained from CentOS6 and CentOS7 are colored in purple and green squares.
    % The red boxes highlight regions with significant differences.}
    An animated version of the comparison is available \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/pfs_t2w_alignment.gif}{here}.}
    \label{fig:fnirt_result}
\end{figure}

@@

However, FreeSurfer results still differ between conditions, due to the
propagation of differences created in PreFreeSurfer. We
observed the effect of this propagation in FreeSurfer results, as shown in
Figure~\ref{fig:tissue_class} for whole-brain segmentations. The
Dice coefficients associated with the 44 regions segmented by
FreeSurfer are shown in Figure~\ref{fig:scatter_plot}, showing that Dice coefficents below 0.9 are 
observed in most regions, and particularly in the smallest ones.
\revised{However, there is no significant correlation between Dice value and region size.}

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{figures/brain_segmentation_mni.png} 
  \caption{Sum of binarized differences between whole-brain FreeSurfer
  segmentations obtained from PreFreeSurfer processings in CentOS6 vs CentOS7
   (N=20). Segmentations were resampled and overlaid to the MNI152 volume
  template. 
  \revised{
  The binary images are created based on the intensity differences between result images in CentOS6 and CentOS7. 
  The voxel is 1 if the intensity value of the corresponding voxel in the images is different, otherwise it is 0.
  Each voxel shows the number of subjects that created different results. 
  % The color bar ranges from 1 to 7 as the minimum and the maximum number of subjects in each voxel.
  Voxels are not colored if the identical results were produced in all 20 subjects.}
  An animated comparison of segmentations obtained for a particular subject is available
\href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/fs_brain_segmentation.gif}
{here}.} 
  \label{fig:tissue_class}
\end{figure}

@@

\revision{
Discuss irreproducibility in 2 parts:
1. code coverage 
2. statistics, see eq 1 in https://ieeexplore.ieee.org/abstract/document/7152570
}

@@

\revision{To evaluate the reproducibility of a pipeline, \toolname needs to execute
it 3 times in order to (1) record a first \reprozip trace, (2) save
transient files, and (3) compare results in the second condition. This is
one more execution than the theoretical minimum of 2. It might be possible
to further reduce this overhead by executing at step (2) only the
processes depending on transient files.}

@@

@@

We presented \toolname, a tool to detect the source of numerical
differences in complex pipelines executed in different computational
conditions. \toolname leverages system-call interception through the
\reprozip tool, and therefore can be applied to the most complex pipelines
without requiring their instrumentation. It is available at
\url{https://github.com/big-data-lab-team/spot} under MIT license.

@@

>>>>

\revised{
The labeled graph can differ depending on the order of executions in which condition we capture transient files 
or execute the pipeline to pinpoint the propagation of differences. Therefore, we label processes in two orders 
of execution, and then aggregate findings so that label processes as non-reproducible (red) if it creates 
different results in either order of executions.}

\subsection{Implementation}
\revised{
\toolname is implemented in Python (>=3.5). In this work we used \toolname version 0.2 and the following version of
the Python package dependencies: NumPy v1.19.0~\cite{oliphant2006guide} and Pandas v1.0.5~\cite{mckinney2011pandas},
for data manipulations, SciPy v1.5.1~\cite{2020SciPy-NMeth} and Scikit-learn v.0.23.1~\cite{pedregosa2011scikit}
for the clustering of provenance graphs, Zss v1.2.0~\cite{zhang1989simple} for tree distances,
\reprozip v1.0.11 for the capture of provenance traces, Docker v17.05~\cite{merkel2014docker} for
the edition of container images, and Boutiques v0.5.25~\cite{glatard2018boutiques} for uniform pipeline executions.

Software users will mostly have to interact with the Boutiques and \reprozip packages.
Boutiques is a flexible description framework for containerized pipelines, required by
the pipelines analyzed in \toolname. It provides a JSON schema to describe inputs, outputs and their dependencies.
Examples, tutorials and usage documentation are available at \url{http://boutiques.github.io}. \reprozip intercepts
system calls to identify the files and processes involved in a pipeline execution.
Before using \toolname, users have to collect \reprozip traces of their pipeline executions.
Examples in the \toolname documentation include \reprozip provenance capture.
More documentation on \reprozip is available at \url{https://www.reprozip.org}.}

@@

We randomly selected 20 unprocessed subjects from the HCP data release S500
available in \href{https://db.humanconnectome.org}{the ConnectomDB
repository} as a subset of the 1200 Subject Release \revised{(see Supplementary Table S1)}.
For each subject, available data consisted of 1 or 2 T1-weighted images and 1 or 2
T2-weighted images, with $256 \times 320 \times 320$ voxels of size $0.7
\times 0.7 \times 0.7$ mm. Acquisition protocols and parameters are
detailed in~\cite{van2013wu}.

@@

% \revision{Using the \reprozip trace obtained from CentOS7,
% we captured transient files in CentOS6 and labeled pipeline processes in CentOS7.}
We processed the 20 subjects with PreFreeSurfer and FreeSurfer, using the 2 CentOS versions.
\revised{First, PreFreeSurfer was utilized to analyze the unprocessed data, and then its results 
from CentOS6 were used as the input of FreeSurfer in both conditions. 
We also used the ReproZip trace file captured in CentOS6 for labeling the processes in both pipelines.}
Each subject was processed twice on the same operating
system to detect within-OS variability coming from pseudo-random
operations. We compared pipeline results using  FreeSurfer tools \texttt{mri\_diff},
\texttt{mris\_diff}, and \texttt{lta\_diff}, to ignore execution-specific information such as file path or
timestamps. To compare segmentations $X$ and $Y$, we used the Dice coefficient defined as follows:
\[DICE=\frac{2|X \cap Y|}{|X| + |Y|}\]

\revised{
The Dice coefficient~\cite{dice1945measures} is a commonly used metric to validate medical image segmentation.
Dice values range from 0 to 1, with 1 indicating a perfect overlap between two segmentation results and 0 indicates no overlap.
Alternatively, the Jaccard coefficient~\cite{jaccard1912distribution} could be used;
there is a direct correspondence between both metrics.}

@@

\section{Results}
\revised{
All experiments were run on a machine with a 3.4GHz Intel Core i7 processor using 8 cores and 32GB of RAM running
the CentOS7 operating system and Linux kernel version 3.10.
The processing time, output file size, number of file accesses and number of processes observed in PreFreeSurfer
and FreeSurfer are shown in~Table~\ref{table:pipeline-stats}.
The scripts and analyses used to create the figures in this section are available at
\url{https://github.com/big-data-lab-team/HCP-reproducibility-paper}.

\begin{table*}[ht]
  \centering
  \begin{threeparttable}
  \caption{Execution statistics of the pipelines per subject.}
  \label{table:pipeline-stats}
  \begin{tabular}{ccc|cc}
  \toprule
  \multicolumn{1}{c}{} & \multicolumn{2}{c}{PreFreeSurfer } &   \multicolumn{2}{c}{FreeSurfer } \\

   &        \makecell{Mean}   &  \makecell{Standard error}  & \makecell{Mean}   &  \makecell{Standard error}   \\ \midrule
  Processing time (mins)      &     106.67     &     2.68     &     nd     &     nd     \\
  Output file size (GB)       &     2.8        &     0.10     &     nd     &     nd     \\
  Number of file accesses     &     94,089      &     2,645  &     nd     &     nd     \\
  Number of processes         &     8731       &     198   &     nd     &     nd     \\
  \bottomrule
  \end{tabular}
  \end{threeparttable}
  \end{table*}
}
% All experiments were run on a machine with a 3.4GHz Intel Core i7 processor
% and 32GB memory under the CentOS7 operating system.
% The average processing time per subject was approximately 2~hours (with standard error X)
% for PreFreeSurfer and 8~hours (with standard error X) for FreeSurfer.
% The average output file size was 2.7~GB (with standard error Y) for PreFreeSurfer
% and 4.1~GB (with standard error Y) for FreeSurfer.
%\tristan{Ali, are these numbers per subject or for 20 subjects? it's an average of 20 subjects}
% For each subject, PreFreeSurfer accessed 83K files and created 7.7K processes, and FreeSurfer accessed 62K files and
% created 4K processes.

@@

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figures/t2w_alignment.png}
    \caption{Differences between T2 \fnirt results in PreFreeSurfer's Brain Extraction (CentOS6 vs CentOS7).
    \revised{
    The colored squares indicate results obtained with CentOS6 (in purple) and CentOS7 (in green).
    The red boxes highlight regions with significant differences between the two OSes. }
    An animated version of the comparison is available \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/pfs_t2w_alignment.gif}{here} for better visualization.}
    \label{fig:fnirt_result}
\end{figure}

@@

However, FreeSurfer results still differ between conditions, due to the
propagation of differences created in PreFreeSurfer. We
observed the effect of this propagation in FreeSurfer results, as shown in
Figure~\ref{fig:tissue_class} for whole-brain segmentations. The
Dice coefficients associated with the 44 regions segmented by
FreeSurfer are shown in Figure~\ref{fig:scatter_plot}, showing that Dice coefficents below 0.9 are
observed in most regions, and particularly in the smallest ones.
\revised{However, no significant correlation between the Dice values and the region sizes was found
(Pearson's coefficient = 0.12, p-value = 0.43).}

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{figures/brain_segmentation_mni.png}
  \caption{Sum of binarized differences between whole-brain FreeSurfer
  segmentations obtained from PreFreeSurfer processings in CentOS6 vs CentOS7
   (N=20). Segmentations were resampled and overlaid to the MNI152 volume
  template.
  \revised{
  Each voxel shows the number of subjects for which different results were observed between CentOS 6 and CentOS 7.}
  An animated comparison of segmentations obtained for a particular subject is available
\href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/fs_brain_segmentation.gif}
{here} for better visualization.}
  \label{fig:tissue_class}
\end{figure}

@@

@@

\revised{
To evaluate the reproducibility of a pipeline, \toolname needs to execute it 5 times 
in order to (1) record a first \reprozip trace, (2) save transient files in the first condition, 
(3) compare results in the second condition, and repeat steps (2) and (3) for the other order of execution. 
It might be possible to further reduce this overhead by executing at step (2) only the processes 
depending on transient files, and capturing the transient files for the second condition simultaneously at step (3).}

@@

\revised{
% Furthermore, we can use methods to ensure that enough tests have been performed for labeling a tool as reproducible.
% We approved the irreproducibility of the tools, but we cannot say a tool is reproducible since
\toolname labels pipeline processes based on the tested dataset, 
while the reproducibility of the tools can be changed over the new datasets. 
In this regard, we need to confirm the reproducibility of the tools over all the possible 
subject diversity and processing parameters. 
We can rely on the code coverage to measure the proportion of the code covered by the tests. 
Ideally, the code coverage as close to 100\% as possible ensures that most of the code is fully tested. 
This is a key principle of reproducibility to guarantee the appropriate testing of the tools. 
Having sufficient subject tests so that covers most of the code could increase confidence to confirm 
the reproducibility of the tools.
As another solution, we can estimate the probability of the tool reproducibility based on 
the number of subjects that the tool was tested as reproducible. To this order, 
we refer to the probability measurement used in~\cite{glatard2015classifications}, which can state that the probability 
to observe all reproducible results among $N’$ future tests given that $n$ non-reproducible results happened among $N$ 
previous tests is:

\[p(N',n,N)=\frac{N+1}{N+N'+1}\frac{\binom Nn}{\binom {N+N'}n}\]

This formula gives a high probability to see reproducible results from the tools have been tested on a high series of 
subjects as reproducible.
}

@@

We presented \toolname, a tool to detect the source of numerical
differences in complex pipelines executed in different computational
conditions. \toolname leverages system-call interception through the
\reprozip tool, and therefore can be applied to the most complex pipelines
without requiring their instrumentation. It is available at
\url{https://github.com/big-data-lab-team/spot} under MIT license.
% Moreover, we registered the spot tool in the bio.tools with the identifier biotools:spottool,
% and in the SciCrunch.org database with RRID (Research Resource Identifier) SCR\_018915.

@@

\revised{
\section{Availability of Source Code and Requirements}
\begin{itemize}
  \item Project name: \toolname
  \item Project home page:\url{https://github.com/big-data-lab-team/spot}
  \item Operating system: Linux
  \item Programming language: Python
  \item Other requirements: Python 3.6 or higher, \reprozip, Docker, and Boutiques
  \item License: MIT License
  \item Biotools identifier: \href{https://bio.tools/spottool}{spottool}
  \item SciCrunch ID: RRID:SCR\_018915

\end{itemize}

\section{Supplementary Data}
Supplementary materials include Supplementary Table S1. Summary of the subjects used in the experiments.
}
